---
title: "Homework 4"
subtitle: <center> <h1>Multiple Linear Regression</h1> </center>
author: <center> Madison Wozniak <center>
output: html_document
---

```{=html}
<style type="text/css">
h1.title {
  font-size: 40px;
  text-align: center;
}
</style>
```
```{r setup, include=FALSE}
library(tidyverse)
library(ggfortify) # needed for autoplot with lm objects
library(car)  # needed for added-variable plots 
library(corrplot)  # colored correlation matrix
library(patchwork)
library(GGally)
```

## Data and Description

Measuring body fat is not simple. One method requires submerging the body underwater in a tank and measuring the increase in water level. A simpler method for estimating body fat would be preferred. In order to develop such a method, researchers recorded age (years), weight (pounds), height (inches), and three body circumference measurements (around the neck, chest, and abdominal (all in centimeters)) for 252 men. Each man's percentage of body fat was accurately estimated by an underwater weighing technique (the variable "brozek" is the percentage of body fat). The hope is to be able to use these data to create a model that will accurately predict body fat percentage using just the basic variables recorded, without having to use the tank submerging method.

The data can be found in the BodyFat data set on Canvas. Download "BodyFat.txt", and put it in the same folder as this R Markdown file.

#### 0. Replace the text "\< PUT YOUR NAME HERE \>" (above next to "author:") with your full name.

#### 1. Read in the data set, and call the data frame "bodyfat". Print a summary of the data. **Remove the "row" column (which contains row numbers) from the data set.**

```{r}
bodyfat <- read_table("BodyFat.txt")
summary(bodyfat)
bodyfat <- bodyfat[-1]
```

#### 2. Create and print a scatterplot matrix of the data.

```{r, fig.align='center'}
#ggpairs(bodyfat, mapping = aes(alpha = 0.3))
plot(bodyfat)
```

#### 3. Based on the scatterplot matrix, briefly explain which variables you think will be "significant" for predicting brozek and which variables you think will *not* be helpful at predicting brozek. Explain how the scatterplot helped determine your answers.

I think weight, chest, and abdominal would be the most significant variables for predicting brozek because they have the strongest positive linear correlation with brozek.

#### 4. Create and print a correlation matrix (numeric or color- and shape-coded).

```{r}
corrplot(cor(bodyfat), type = "upper")
```

#### 5. Based on the scatterplot matrix and the correlation matrix, are there any pairs of variables that you suspect will cause a problem in terms of multicollinearity? If so, which ones?

chest/abdominal, weight/chest, weight/abdominal, and weight/neck have the highest correlations with one another.

#### 6. Fit a multiple linear regression model to the data (no transformations). Print a summary of the results. Save the residuals to the `bodyfat` data frame.

```{r}
bodyfat.lm <- lm(brozek ~ ., data = bodyfat)
summary(bodyfat.lm)
bodyfat$residuals <- bodyfat.lm$residuals
bodyfat$fitted_vals <- bodyfat.lm$fitted.values
```

#### 7. Briefly comment on the "significance" of the variables: were you surprised by the results? Are there any variables that are significant that you think shouldn't be? Are there any variables that are not significant that you think should be?

weight, neck, and abdominal are statistically significant variables, which isn't surprising because looking at the scatterplot matrix I was able to infer that these might have the highest correlation with brozek.

#### 8. Briefly comment on the sign (+/-) of the coefficients for the variables. Are their any variables where the sign is the opposite of what you expected?

Negative coefficients suggest that an increase in that variable have a negative effect on body fat (brozek). I would expect that an increase in weight would have the opposite effect on body fat, as well as height.

#### 9. Mathematically write out the *fitted* multiple linear regression model for this data set using the coefficients you found above (do not use betas). Do not use "X" and "Y" in your model - use variable names that are fairly descriptive.

$brozek_i$ = $-2.010e+01$ + 5.010e-03 \* $age_i$ - 8.733e-02 \* $weight_i$ - 1.400e-01 \* $height_i$ - 4.421e-01 \* $neck_i$ + 4.844e-04 \* $chest_i$ + 8.754e-01 \* $abdominal_i$

#### 10. *Assuming* the model assumptions are all met, how would you interpret the coefficient for Weight?

Holding all other predictors constant, for every one pound increase in weight, the average brozek measurement will decrease by 8.733e-02.

#### 11. Briefly explain what it means to "hold all else constant" when interpreting a coefficient.

This means we want to keep all the predictor variables we are not currently looking at constant, and only comparing the conditional relationship between X and Y (in our case brozek and Weight).

#### 12. Briefly explain what the F-test indicates, as given in the model output from question 6.

The F-test tells us whether or not at least one predictor variable is significantly linearly correlated with the response variable. Since our F-test returns a very low p-value and high F-statistic, we can conclude that there is at least one variable that is linearly associated with average Brozek.

#### 13. Briefly interpret the *adjusted* R-squared, as reported in the model output from question 6.

About 72% of the total variation in body fat percentage is explained by at least one of the predictors in the model, after accounting for the number of variables in the model.

### Questions 14-20 involve using diagnostics to determine if the linear regression assumptions are met. For each assumption, (1) perform appropriate diagnostics to determine if the assumption is violated, and (2) explain whether or not you think the assumption is violated and why you think that.

#### 14. The X's vs Y are linear (use the residual vs. predictor plots, partial regression plots, and the residual vs. fitted values plot).

```{r, fig.align='center'}
# residual vs. predictor plots
resid_vs_age <- ggplot(data = bodyfat) +
  geom_point(mapping = aes(x = age, y = residuals)) +
  theme(aspect.ratio = 1)
resid_vs_weight <- ggplot(data = bodyfat) +
  geom_point(mapping = aes(x = weight, y = residuals)) +
  theme(aspect.ratio = 1)
resid_vs_height <- ggplot(data = bodyfat) +
  geom_point(mapping = aes(x = height, y = residuals)) +
  theme(aspect.ratio = 1)
resid_vs_neck <- ggplot(data = bodyfat) +
  geom_point(mapping = aes(x = neck, y = residuals)) +
  theme(aspect.ratio = 1)
resid_vs_chest <- ggplot(data = bodyfat) +
  geom_point(mapping = aes(x = chest, y = residuals)) +
  theme(aspect.ratio = 1)
resid_vs_abdom <- ggplot(data = bodyfat) +
  geom_point(mapping = aes(x = abdom, y = residuals)) +
  theme(aspect.ratio = 1)
(resid_vs_age | resid_vs_weight | resid_vs_height) /
  (resid_vs_neck | resid_vs_chest | resid_vs_abdom)
```

```{r, fig.align='center'}
# partial regression plots
avPlots(bodyfat.lm)
```

```{r, fig.align='center'}
# residual vs fitted values
autoplot(bodyfat.lm, which = 1, ncol = 1, nrow = 1) +
  theme(aspect.ratio = 1)
```

All three plots suggest that the residuals are mostly linear, the Residuals vs Fitted model is pulled down toward the end, but I suspect this is due to an influential point which does not reflect the overall behavior of the residuals.

#### 15. The residuals are independent (I will answer this one for you, no need to modify response. No points for this question).

Since we do not know if the 252 men were randomly sampled from a population, we do not know if the residuals are independent or not. We will assume that they are independent for this analysis.

#### 16. The residuals are normally distributed (use a histogram, qq-plot, and Shapiro-Wilk test)

```{r, fig.align='center'}
# Diagnostic 1
ggplot(data = bodyfat, mapping = aes(y = residuals)) + 
  geom_boxplot() + 
  scale_x_discrete() + 
  labs(title = "Residual Boxplot")
```

```{r, fig.align='center'}
# Diagnostic 2
qqPlot(bodyfat.lm)
```

```{r, fig.align='center'}
# Diagnostic 3
shapiro.test(bodyfat$residuals)
```

All three diagnostics used tell us that the residuals are normally distributed.

#### 17. The residuals have equal/constant variance across all values of X (check Scale-location plot and residuals vs. fitted values plot)

```{r, fig.align='center'}
autoplot(bodyfat.lm, which = 3, ncol = 1, nrow = 1) 
```

```{r, fig.align='center'}
autoplot(bodyfat.lm, which = 1, ncol = 1, nrow = 1) +
  theme(aspect.ratio = 1)
```

Based on the two diagnostic tests above, I would conclude that the residuals have equal variance, of course, aside from our influential point.

#### 18. Check for no influential points using Cook's distance. Also, in your response, reference the plots you created in previous questions.

```{r, fig.align='center'}
# Cook's Distance
autoplot(bodyfat.lm, which = 4, ncol = 1, nrow = 1) 
```

There is one influential point, which is observation number 39, which makes sense because in the Residuals vs Fitted Values plot, and Scale-Location plot above show that observation 39 is an outlier as well.

#### 19. Check for extreme multicollinearity. For this (tacit) model assumption, compute the variance inflation factors (VIFs) and compare the VIFs to your response in question 5. Is there agreement? Is this assumption met (recall: the rule of thumb is that each VIF should be less than 10 and the average of the VIFs should be close to 1)?

```{r, fig.align='center'}
vif(bodyfat.lm)
```

The predictor variables that have extreme multicollinearity are Weight and abdominal. In question 5 I predicted chest would also have extreme multicollinearity, but this output shows that it is not as severe as weight and abdominal.

### Note: your next homework assigment will use this same data set, and you will be asked to fix the assumptions that were broken.

#### 20. Briefly summarize what you learned, personally, from this analysis about the statistics, model fitting process, etc.

From this analysis about statistics and model fitting, I learned how useful matrix scatterplots can be when getting a general idea of the correlation between variables is necessary. I also found the correlation matrix helpful because it gave a new way of visualizing how large correlations are.

#### 21. Briefly summarize what you learned from this analysis *to a non-statistician*. Write a few sentences about (1) the purpose of this data set and analysis and (2) what you learned about this data set from your analysis. Write your response as if you were addressing a business manager (avoid using statistics jargon) and just provide the main take-aways.

The purpose of our dataset is to try to find a linear relationship between brozek (body fat percentage) and a handful of variables including age, weight, height, neck length, chest size, and abdominal. What I learned from conducting an analysis on the data provided, is that weight, neck, and abdominal have the highest relationship with body fat percentage. On average, an increase in body fat percent is associated with higher abdominal size, and a decrease in neck and weight result in a decrease in body fat percentage.
